---
title: "Introduction to Bayesian linear regression with brms"
author: "Stefano Coretta"
date: "18/01/2020"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: zenburn
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(brms)
library(coretta2018itaegg)
data("ita_egg")
```

## Random variables

* We have a question about the world, so we collect data (sample from a population).
  * y = ($y_1$, $y_2$, $y_3$, $y_4$, ..., $y_n$)
* We want to know how the data (the sample) was generated.
* In probability theory, data is generated by a random variable $Y$.

## Random variables

* $Y$ is uncertain.
  * We can describe $Y$ as a probability distribution, expressed by a set of parameters $\Theta = (\theta_1, ..., \theta_n)$.
* Probability distributions:
  * $Normal(\mu, \sigma)$,
  * $Binomial(n, p)$,
  * ...

## Random variables

\centering \Large

$vot \sim Normal(\mu, \sigma)$

$voiced \sim Bernoulli(p)$

$DoubleDative \sim Poisson(\lambda)$

## Frequentist vs Bayesian view

* Parameters: $\mu$, $\sigma$, $p$, $\lambda$, ...
* Frequentist view:
  * The parameters are **fixed** (they are unknown but certain).
  * They take on a specific value.
* Bayesian view:
  * The parameters are **random variables** (they are unkown and uncertain).
  * We describe each parameter as a probability distribution, expressed by a set of **hyperparameters**.

## Continous random variable

\centering \Large

$vot_i \sim Normal(\mu, \sigma)$

$\mu \sim Normal(\mu_1, \sigma_1)$

$\sigma \sim HalfCauchy(x_0, \gamma)$

## Priors

* We can incorporate previous knowledge about the hyperparameters as **priors** (prior distributions).
* Priors are chosen based on expert knowledge, previous studies, pilot data...
  * Priors must **not** be chosen based on the data to be analysed.

## Priors

* Informative and uninformative priors.
* Informative priors.
  * Weakly informative.
  * Strongly informative.
* Uninformative priors.
  * Uniform distribution.

## Normal prior

## Italian VOT

* Previous literature on VOT in Italian \citep{esposito2002, stevens2010a} report VOT values for voiceless stops in the range of 20--60 ms.
  * We can express this knowledge with the prior $Normal(40, 10)$.
  * This is a somewhat strongly informative prior.

## Italian VOT

```{r mean-prior}
plot(-30:110, dnorm(-30:110, 40, 10), type = "l", ann = FALSE)
```

## Italian VOT

```{r mean-prior-2}
plot(-30:110, dnorm(-30:110, 40, 20), type = "l", ann = FALSE)
```

## Italian VOT


