---
title: "Introduction to Bayesian linear regression with brms"
author: "Stefano Coretta"
date: "18/01/2020"
output:
  beamer_presentation:
    citation_package: natbib
    highlight: tango
    latex_engine: xelatex
fontsize: 12pt
bibliography: linguistics.bib
biblio-style: unified.bst
header-includes:
- \frenchspacing
- \usepackage{cleveref}
- \usetheme{metropolis}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(brms)
options(mc.cores = parallel::detectCores())
library(extraDistr)
library(coretta2018itaegg)
data("ita_egg")
ita_egg <- drop_na(ita_egg, vot)
```

```{r read-bernoulli-data, message=FALSE}
pilot <- list.files(
  path = "./data/perceptual/",
  pattern = "*.csv",
  full.names = TRUE
) %>%
  map_df(~read_csv(., col_types = cols(.default = "c"))) %>%
  mutate(
    key_resp_2.keys = ifelse(key_resp_2.keys == "z", "d", "t"),
    key_resp_8.keys = ifelse(key_resp_8.keys == "z", "t", "d"),
    response = coalesce(key_resp_2.keys, key_resp_8.keys),
    response = factor(response, levels = c("t", "d")),
    burst = as.numeric(burst),
    voicing = as.numeric(voicing),
    response_n = as.numeric(response) - 1,
    vowel = factor(vowel, levels = c("a", "i", "u"))
  ) %>%
  filter(!is.na(response))

contrasts(pilot$vowel) <- "contr.sum"

burst <- filter(pilot, condition == "burst")
voicing <- filter(pilot, condition == "voicing")
```

## Installation

[mention installation]

## Random variables

* We have a question about the world, so we collect data (sample from a population).
  * y = ($y_1$, $y_2$, $y_3$, $y_4$, ..., $y_n$)
* We want to know how the data (the sample) was generated.
* In probability theory, data is generated by a random variable $Y$.

## Random variables

* $Y$ is uncertain.
  * We can describe $Y$ as a probability distribution, expressed by a set of parameters $\Theta = (\theta_1, ..., \theta_n)$.
* Probability distributions:
  * $Normal(\mu, \sigma)$,
  * $Binomial(n, p)$,
  * ...

## Random variables

\centering \Large

$vot_i \sim Normal(\mu, \sigma)$

$voiced_i \sim Bernoulli(p)$

$DoubleDative_i \sim Poisson(\lambda)$

## Frequentist vs Bayesian view

* Parameters: $\mu$, $\sigma$, $p$, $\lambda$, ...
* Frequentist view:
  * The parameters are **fixed** (they are unknown but certain).
  * They take on a specific value.
* Bayesian view:
  * The parameters are **random variables** (they are unkown and uncertain).
  * We describe each parameter as a probability distribution, expressed by a set of **hyperparameters**.

## Continous random variable

\centering \Large

$vot_i \sim Normal(\mu, \sigma)$

$\mu \sim Normal(\mu_1, \sigma_1)$

$\sigma \sim HalfCauchy(x_0, \gamma)$

## Bayes' Theorem

$$ P(\theta \mid d) = \frac{P(d \mid \theta) \, P(\theta)}{P(d)} $$

## Bayes' Theorem

$$ posterior\ probability = \frac{likelihood \times prior}{marginal\ likelihood} $$

## Priors

* We can incorporate previous knowledge about the hyperparameters as **priors** (prior distributions).
* Priors are chosen based on expert knowledge, previous studies, pilot data...
  * Priors must **not** be chosen based on the data to be analysed.

## Priors

* Informative and weakly informative priors.
* Uninformative or diffuse priors.
  * Uniform distribution.
* Regularising priors.

## Normal prior

[empirical rule]

## Italian VOT

* Previous literature on VOT in Italian \citep{esposito2002, stevens2010a} report VOT values for voiceless stops in the range of 20--60 ms.
  * We can express this knowledge with the prior $Normal(40, 10)$.
  * This is a somewhat strongly informative prior.

## Italian VOT

```{r mean-prior}
plot(-30:110, dnorm(-30:110, 40, 10), type = "l", main = "Normal(40, 10)", xlab = "", ylab = "")
```

## Italian VOT

```{r mean-prior-2}
plot(-30:110, dnorm(-30:110, 40, 20), type = "l", main = "Normal(40, 20)", xlab = "", ylab = "")
```

## Italian VOT

\centering \Large

$vot_i \sim Normal(\mu, \sigma)$

$\mu \sim Normal(40, 10)$

$\sigma \sim HalfCauchy(x_0, \gamma)$

## Cauchy prior

```{r cauchy-prior}
plot(seq(-100, 100, by = 0.1), dcauchy(seq(-100, 100, by = 0.1), 0, 10), type = "l", main = "Cauchy", xlab = "", ylab = "")
lines(seq(-100, 100, by = 0.1), dcauchy(seq(-100, 100, by = 0.1), 0, 20), type = "l")
lines(seq(-100, 100, by = 0.1), dcauchy(seq(-100, 100, by = 0.1), 0, 30), type = "l")
```

## Cauchy prior

```{r cauchy-prior-2}
plot(seq(0, 100, by = 0.1), dcauchy(seq(0, 100, by = 0.1), 0, 10), type = "l", main = "HalfCauchy", xlab = "", ylab = "")
lines(seq(0, 100, by = 0.1), dcauchy(seq(0, 100, by = 0.1), 0, 20), type = "l")
lines(seq(0, 100, by = 0.1), dcauchy(seq(0, 100, by = 0.1), 0, 30), type = "l")
```

## Italian VOT

\centering \Large

$vot_i \sim Normal(\mu, \sigma)$

$\mu \sim Normal(40, 10)$

$\sigma \sim HalfCauchy(0, 10)$

## Italian VOT

* We have a model which incorporates (some of) our knowledge about VOT (through the priors for $\mu$ and $\sigma$).
* Now we want to obtain the **posterior distributions** of $\mu$ and $\sigma$.
  * The posterior distribution is the prior distribution *conditioned* on the data.
* **brms** R package: Bayesian Regression Models using Stan \citep{burkner2018}.

## brms

* Stan \citep{stan-development-team2017}.
  * Statistical programming language written in C++ for fitting Bayesian models (calculate posterior distributions).
  * Calculation can be complex and/or impossible, so we take many samples from the data and from the possible parameter values to find the posterior distributions of the hyperparameters.
  * Markov Chain Monte Carlo (MCMC) sampling using the No-U-Turn sampler (NUTS).
* brms is an interface between R and Stan.
* `brm()` function from brms.
  * lme4 syntax (`y ~ x + (1|w)`).
  * Creates a Stan model, which is compiled and run.

## brms

```{r brm, eval=FALSE, echo=TRUE}
library(brms)

vot1 <- brm(
  <model_formula>,
  <family>,
  <prior>,
  <data>,
  chains = 4,
  iter = 2000
)
```

## brms

```{r brm-2, eval=FALSE, echo=TRUE}
library(brms)

vot1 <- brm(
  vot ~ 1,
  family = gaussian(),
  <prior>,
  data = ita_egg,
  chains = 4,
  iter = 2000
)
```

## Get prior

```{r get-prior, echo=TRUE}
get_prior(
  vot ~ 1,
  family = gaussian(),
  data = ita_egg
)
```

## Set prior

```{r set-prior}
priors <- c(
  prior(normal(40, 10), class = Intercept),
  prior(cauchy(0, 10), class = sigma)
)
```

## Prior predictive checks

```{r vot1-priors, echo=TRUE}
nsim <- 1000
nobs <- 100

y <- matrix(rep(NA, nsim * nobs), ncol = nobs)

mu <- rnorm(nsim, 40, 10)
sigma <- rhcauchy(nsim, 10)

for (i in 1:nsim) {
  y[i,] <- rnorm(nobs, mean = mu[i], sd = sigma[i])
}
```

## Prior predictive checks

```{r vot1-priors-plot, echo=TRUE}
op <- par(mfrow = c(3, 3))
rand_sample <- sample(1:nsim, 9)

for (i in rand_sample) {
  plot(density(y[i,]), main = "")
}

par(op)
```


## Run the model

```{r vot1, echo=TRUE}
vot1 <- brm(
  vot ~ 1,
  family = gaussian(),
  prior = priors,
  data = ita_egg,
  chains = 4,
  iter = 2000,
  file = "./cache/vot1"
)
```

## Model summary

```{r vot1-sum, echo=TRUE}
vot1
```

## Plot model

```{r vot1-plot, echo=TRUE}
plot(vot1)
```

## Posterior predictive check

```{r vot1-pp, echo=TRUE}
pp_check(vot1, nsamples = 100)
```

## Adding predictors

$vot_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \alpha + \beta_1 \times coronal_i + \beta_2 \times velar_i$

$\alpha \sim Normal(\mu_1, \sigma_1)$

$\beta_1 \sim Normal(\mu_2, \sigma_2)$

$\beta_2 \sim Normal(\mu_3, \sigma_3)$

$\sigma \sim HalfCauchy(x_0, \gamma)$

## Adding predictors

$vot_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \alpha + \beta_1 \times coronal_i + \beta_2 \times velar_i$

$\alpha \sim Normal(25, 10)$

$\beta_1 \sim Normal(10, 10)$

$\beta_2 \sim Normal(20, 10)$

$\sigma \sim HalfCauchy(0, 10)$

## Adding predictors

```{r get-prior-2, echo=TRUE}
get_prior(
  vot ~ 1 + c1_place,
  family = gaussian(),
  data = ita_egg
)
```

## Adding predictors

```{r set-prior-2, echo=TRUE}
priors <- c(
  prior(normal(25, 10), class = Intercept),
  prior(cauchy(0, 10), class = sigma),
  prior(normal(10, 10), class = b, coef = "c1_placecoronal"),
  prior(normal(20, 10), class = b, coef = "c1_placevelar")
)
```

## Adding predictors

```{r vot2, echo=TRUE}
vot2 <- brm(
  vot ~ 1 + c1_place,
  family = gaussian(),
  prior = priors,
  data = ita_egg,
  chains = 4,
  iter = 2000,
  file = "./cache/vot2"
)
```

## Random effects

\centering \small

$vot_i \sim Normal(\mu_i, \sigma)$

$\mu_i = \alpha + \alpha_{speaker[i]} + (\beta_1 + \beta_{1speaker[i]}) \times coronal_i + (\beta_2 + \beta_{2speaker[i]}) \times velar_i$

$$\begin{bmatrix}
\alpha_{speaker}\\
\beta_{1speaker[i]}\\
\beta_{2speaker[i]}
\end{bmatrix} \sim MVNormal(\begin{bmatrix}0\\0\\0\end{bmatrix}, S)$$

$\alpha \sim Normal(25, 10)$

$\alpha_{speaker} \sim Normal(0, \sigma_{speaker})$

$\beta_1 \sim Normal(10, 10)$

$\beta_2 \sim Normal(20, 10)$

$\sigma_{\alpha{speaker}} \sim Normal(0, \sigma_{speaker})$

$\sigma_{\beta{1speaker}} \sim HalfCauchy(0, 10)$

$\sigma_{\beta{2speaker}} \sim HalfCauchy(0, 10)$

$\sigma_{speaker} \sim HalfCauchy(0, 10)$

$\sigma \sim HalfCauchy(0, 10)$

## Random effects

```{r get-prior-3, echo=TRUE}
get_prior(
  vot ~ 1 + c1_place + (1 + c1_place | speaker),
  family = gaussian(),
  data = ita_egg
)
```

## Random effects

```{r set-prior-3, echo=TRUE}
priors <- c(
  prior(normal(40, 10), class = Intercept),
  prior(cauchy(0, 10), class = sigma),
  prior(normal(10, 10), class = b, coef = "c1_placecoronal"),
  prior(normal(20, 10), class = b, coef = "c1_placevelar"),
  prior(normal(0, 25), class = sd),
  prior(lkj(2), class = cor)
)
```

## Random effects

```{r vot3, echo=TRUE}
vot3 <- brm(
  vot ~ 1 + c1_place + (1 + c1_place | speaker),
  family = gaussian(),
  prior = priors,
  data = ita_egg,
  chains = 4,
  iter = 2000,
  file = "./cache/vot3"
)
```

## Random effects

```{r vot3-plot}
plot(vot3)
```

## Random effects

```{r vot3-sum, echo=TRUE}
vot3
```

## Binomial logistic regression

```{r set-prior-burst, echo=TRUE}
priors <- c(
  prior(student_t(3, 0, 2), class = Intercept),
  prior(student_t(3, 0, 2), class = b),
  prior(cauchy(0, 1), class = sd),
  prior(lkj(2), class = cor)
)
```

## Binomial logistic regression

```{r burst, echo=TRUE}
burst1 <- brm(
  response_n ~
    burst *
    vowel +
    (1+burst|participant),
  data = burst,
  prior = priors,
  family = bernoulli,
  file = "./cache/burst1",
  control = list(adapt_delta = 0.999)
)
```

## Binomial logistic regression

```{r plot-cond-burst, echo=TRUE}
conditional_effects(burst1, effects = "burst:vowel")
```

